Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2025-02-04 report/emse24-revised.tex
chore(report): emse24-revised updates

\section{Introduction}

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for all tokens is calculated and then normalized using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value. We apply stratified random sampling to identify the candidates for the case study analysis. The subgroups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value. The candidates are then divided into quartiles based on the aggregate value. A candidate is randomly drawn from each bin and analysed as an in-depth case study. The analysis is stopped when the time resource is exhausted.

During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Further discussions were held with all authors until a consensus was reached.

@@

The results of the study are presented in this section. For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. This is done only to structure the paper in a meaningful manner and for ease of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples. These can be viewed on our online appendix\footnote{https://arumoy.me/shome2023notebook}.

@@

\highlight{\textbf{Finding 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion. 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries. Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}

@@

\highlight{\textbf{Finding 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message. These assertions frequently perform a comparison check or call an external function. The comparison checks frequently verify that the shape and size of data structures is equal to a specified value.}

@@

\highlight{\textbf{Finding 1.3} Print statements are most commonly used to display various performance metrics of ML models.}

@@

\highlight{\textbf{Finding 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}

@@

\highlight{\textbf{Finding 1.5} Common functions called on pandas dataframes focus on exploratory data analysis, statistical operations and quick visualizations. In visualizations, functions for categorical data, correlation and distribution are frequently used.}

@@

\highlight{\textbf{Finding 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings. These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}

@@

\highlight{\textbf{Finding 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values. These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

@@

\highlight{\textbf{Finding 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics. These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}

@@

\highlight{\textbf{Finding 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values. These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}

@@

\highlight{\textbf{Finding 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes. These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}

@@

\highlight{\textbf{Finding 2.6} Five assertions validate the data types of input features, models, and intermediate objects. These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}

@@

\highlight{\textbf{Finding 2.7} Four assertions validate key mathematical properties of data and model outputs. These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry. These checks confirm the robustness and reliability of model computations.}

@@

\highlight{\textbf{Finding 2.8} This study identifies three assertions validating batch size configurations. These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size. These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}

@@

\highlight{\textbf{Finding 2.9} Three assertions verify the consistency of neural network architecture. These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected. These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}

@@

\highlight{\textbf{Finding 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage. Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data. This practice maintains the integrity and validity of the model evaluation metrics.}

@@

\highlight{\textbf{Finding 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models. These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples. These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}

@@

\highlight{\textbf{Finding 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes. These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}

@@

\highlight{\textbf{Finding 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed. These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions. These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}

@@

\highlight{\textbf{Finding 3.4} This study finds five implicit spot checks performed at various stages of the MLDL. Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}

@@

\highlight{\textbf{Finding 3.5} The study identifies four implicit checks used to monitor model training progress. These checks provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods. These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}

@@

\highlight{\textbf{Finding 3.6} The study identifies three implicit checks for missing data. Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations. Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}

@@

\highlight{\textbf{Finding 3.7} The study identifies three implicit checks used to validate the shape of data. These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}

@@

\highlight{\textbf{Finding 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data. Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

\highlight{\textbf{Finding 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset. These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training. Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}

@@

\highlight{\textbf{Finding 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model. The print statement can be used to evaluate the efficiency of model training, especially when multiple models are trained iteratively.}

@@

\highlight{\textbf{Finding 3.11} The study identifies one print statement used to show the architecture of a neural network. Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}

@@

The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged. This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\citep{wang2020assessing}.

@@

This study further reveals several avenues for research. We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them. Further research is required to explore the feedback mechanisms in more detail and organize the knowledge into a taxonomy based on specific ML validation tasks. Future work can focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.

@@

@@

>>>>

\section{Introduction}
The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \lstinline{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.

% TODO: update this paragraph, candidates were annotated by first and second author
% we calculated IRA and obtained high agreement...
During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Further discussions were held with all authors until a consensus was reached.

@@

Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples. These can be viewed on our online appendix\footnote{https://arumoy.me/shome2023notebook}.

Due to the exploratory nature of this study and the limited number of case studies analyzed, the observations presented should be interpreted with caution and viewed as illustrative rather than definitive. The case studies serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.

@@

\highlight{\textbf{Observation 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion. 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries. Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}

@@

\highlight{\textbf{Observation 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message. These assertions frequently perform a comparison check or call an external function. The comparison checks frequently verify that the shape and size of data structures is equal to a specified value.}

@@

\highlight{\textbf{Observation 1.3} Print statements are most commonly used to display various performance metrics of ML models.}

@@

\highlight{\textbf{Observation 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}

@@

\highlight{\textbf{Observation 1.5} Common functions called on pandas dataframes focus on exploratory data analysis, statistical operations and quick visualizations. In visualizations, functions for categorical data, correlation and distribution are frequently used.}

@@

\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings. These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}

@@

\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values. These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

@@

\highlight{\textbf{Observation 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics. These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}

@@

\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values. These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}

@@

\highlight{\textbf{Observation 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes. These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}

@@

\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features, models, and intermediate objects. These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}

@@

\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs. These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry. These checks confirm the robustness and reliability of model computations.}

@@

\highlight{\textbf{Observation 2.8} This study identifies three assertions validating batch size configurations. These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size. These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}

@@

\highlight{\textbf{Observation 2.9} Three assertions verify the consistency of neural network architecture. These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected. These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}

@@

\highlight{\textbf{Observation 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage. Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data. This practice maintains the integrity and validity of the model evaluation metrics.}

@@

\highlight{\textbf{Observation 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models. These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples. These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}

@@

\highlight{\textbf{Observation 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes. These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}

@@

\highlight{\textbf{Observation 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed. These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions. These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}

@@

\highlight{\textbf{Observation 3.4} This study finds five implicit spot checks performed at various stages of the MLDL. Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}

@@

\highlight{\textbf{Observation 3.5} The study identifies four implicit checks used to monitor model training progress. These checks provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods. These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}

@@

\highlight{\textbf{Observation 3.6} The study identifies three implicit checks for missing data. Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations. Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}

@@

\highlight{\textbf{Observation 3.7} The study identifies three implicit checks used to validate the shape of data. These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}

@@

\highlight{\textbf{Observation 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data. Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

\highlight{\textbf{Observation 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset. These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training. Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}

@@

\highlight{\textbf{Observation 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model. The print statement can be used to evaluate the efficiency of model training, especially when multiple models are trained iteratively.}

@@

\highlight{\textbf{Observation 3.11} The study identifies one print statement used to show the architecture of a neural network. Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}

@@

The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged. This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Future work can focus on providing empirical evidence to demonstrate that the integration of assertions significantly improves the reliability of ML workflows in practice.

@@

This study further reveals several avenues for future research. We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them. Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks. It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering. Each task necessitates distinct feedback mechanisms. For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing. By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.

Additionally, future work can also focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.

@@

\subsection{Reproducibility of Jupyter Notebooks}

Code written within Jupyter notebooks can reach a convoluted state since the code cells can be executed in any order. Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them. Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments~\cite{pimentel2019large-scale,wang2020assessing}.

Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells~\citep{wang2020assessing}. Conversely, constructs like \lstinline{print("Accuracy:", accuracy)} are only effective if the cell defining \lstinline{accuracy} has been executed prior to the print statement. In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear. Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.

However, there exists some ambiguity regarding how assertions interact with out-of-order execution in Jupyter notebooks. On one hand, our findings indicate that assertions can effectively prevent the system from reaching a convoluted state that may arise due to such execution.

@@

Furthermore, we would like to reiterate that the case studies presented in RQ2 and RQ3 serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
